---
title: "Data Visualisation Assignment"
author: "Josh Budge"
format: 
  html:
    toc: true
    code-fold: true
---
**LAY READER VERSION**

Do Smaller Animals Really Sleep More?

Some mammals spend much of their day asleep, while others get by on only a few hours. Why such a difference? Scientists have proposed several benefits of sleep, including saving energy and supporting learning and memory. But how long an animal sleeps may also depend on how much time it needs to find food and how safely it can rest without encountering predators (Capellini et al., 2008). Could an animal’s size help explain its sleeping habits?

We explored this question using sleep and body-weight records for 83 mammal species from the msleep dataset. These measurements come from data compiled by Savage and West (2007). Each grey dot in Figure 1 represents one species, showing how many hours it sleeps per day.

We divided the species into two categories using the middle body weight in the dataset: 1.67 kg. The smaller category contains 42 species weighing up to and including 1.67 kg, and the larger category contains 41 species weighing more than 1.67 kg. The teal diamonds show the average sleep duration in each category. Dots are spread slightly sideways to make overlapping species easier to see; their sideways positions do not represent their exact weights.

```{r - layplot1}
#| message: false
# Visualization 1: body size and daily sleep in mammals
# Data: ggplot2::msleep
# Source: https://ggplot2.tidyverse.org/reference/msleep.html
# Sleep and weight data: Savage & West (2007), doi:10.1073/pnas.0610080104


#calling msleep function from package
library(ggplot2)
library(dplyr)

sleep_data <- ggplot2::msleep

# Divide species at the median (middle) body weight.
# factor() keeps "Smaller" before "Larger" on the x-axis.
lay_all <- sleep_data |>
  mutate(
    size_group = if_else(bodywt <= median(bodywt), "Smaller", "Larger"),
    size_group = factor(size_group, levels = c("Smaller", "Larger"))
  )


sleep_plot_1_public <- ggplot(
  lay_all, aes(x = size_group, y = sleep_total)
) +
  # Spread dots sideways only: sleep values stay unchanged.
  # seed makes the sideways positions repeatable when rendering.
  geom_point(
    position = position_jitter(width = 0.12, height = 0, seed = 123),
    size = 1.4, alpha = 0.65, colour = "grey30"
  ) +
  # Calculate the mean sleep for each category and draw a mean bar.
   stat_summary(
    fun = mean,
    geom = "bar",
    width = 0.6,
    fill = "#007C83",
    alpha = 0.25
  ) +
    # Mark the mean with a horizontal teal line.
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 95,
    size = 20,
    colour = "#007C83"
  ) +
  scale_y_continuous(breaks = seq(0, 24, 4)) +
  coord_cartesian(ylim = c(0, 24)) +
  labs(
    title = "Smaller mammals sleep longer on average",
    x = "Body size category",
    y = "Hours of sleep per day",
    caption = "Grey dots: individual species. Teal bars: average sleep."
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold", size = 18),
    plot.caption = element_text(hjust = 0, size = 10)
  )

print(sleep_plot_1_public)
```

The first graph reveals a potentially surprising pattern: smaller mammals tend to sleep longer than larger mammals. The smaller category averages about 12.5 hours of sleep per day, compared with 8.3 hours in the larger category. The grey dots also show that sleep duration varies considerably within each category.

But this broad comparison brings together very different animals. Does the pattern hold when we compare species that are more closely related?

To take a closer look, we split the data into groups of related mammals. We included only groups with at least six species in the dataset, avoiding comparisons based on just a handful of points. This left four groups: rodents such as mice and squirrels (22 species); cats, dogs, seals and their relatives (12 species); primates, including monkeys, apes and humans (12 species); and cattle, deer, pigs and their relatives (6 species).

Together, these groups include 52 of the original 83 species. Within each group, we divided species into smaller and larger halves using that group’s middle body weight. This means “smaller” and “larger” are relative to each group: a larger rodent could still weigh less than a smaller primate. Each panel in Figure 2 shows the individual species and their category averages, with the same sleep scale so they can be compared directly.


```{r - layplot2}
#| message: false


library(ggplot2)
library(dplyr)

# Keep orders with at least six species.
# Calculate a separate median body weight WITHIN each order.
lay_by_order <- sleep_data |>
  add_count(order, name = "n") |>
  filter(n >= 6) |>
  group_by(order) |>
  mutate(
    size_group = if_else(bodywt <= median(bodywt), "Smaller", "Larger")
  ) |>
  ungroup() |>
  mutate(
    size_group = factor(size_group, levels = c("Smaller", "Larger")),
    group_name = case_when(
      order == "Artiodactyla" ~ "Cattle, deer, pigs and relatives",
      order == "Carnivora" ~ "Cats, dogs, seals and relatives",
      order == "Primates" ~ "Monkeys, apes and humans",
      order == "Rodentia" ~ "Mice, squirrels and other rodents",
      TRUE ~ order
    ),
    order_label = paste0(group_name, "\n", n, " species")
  )

sleep_plot_2_public <- ggplot(
  lay_by_order, aes(x = size_group, y = sleep_total)
) +
  geom_point(
    position = position_jitter(width = 0.12, height = 0, seed = 123),
    size = 1.4, alpha = 0.65, colour = "grey30"
  ) +
  # Compute a separate category average within each panel.
  stat_summary(
    fun = mean, geom = "bar",
    fill = "#007C83", alpha = 0.25, colour = "white"
  ) +
    # Mark the mean with a horizontal teal line.
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 95,
    size = 10,
    colour = "#007C83"
  ) +
  facet_wrap(~ order_label, ncol = 2, scales = "fixed", axes = "all_x", axis.labels = "all_x") +
  scale_y_continuous(breaks = seq(0, 24, 4)) +
  coord_cartesian(ylim = c(0, 24)) +
  labs(
    title = "The size difference is less consistent within groups",
    x = "Body size relative to others in the same animal group",
    y = "Hours of sleep per day",
    caption = "Grey dots: individual species. Teal bar: average sleep."
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 18),
    plot.caption = element_text(hjust = 0, size = 10),
    panel.spacing = grid::unit(1.5, "lines")
  )

print(sleep_plot_2_public)
```

The closer look reveals a less consistent pattern. Among rodents and the group containing cats, dogs and seals, the average sleep durations are similar for smaller and larger species. Among primates, however, the larger species average less sleep. For cattle, deer, pigs and their relatives, the larger species actually average more sleep, but there are only three species in each size category, so this comparison should be interpreted cautiously. Dividing body weight into two categories simplifies the comparison but leaves out some of the detail shown in the scientific version’s scatterplots.

The overall pattern therefore tells only part of the story. Smaller mammals generally sleep longer in this dataset, but the relationship becomes less consistent when we compare more closely related species. Some of the broad pattern may reflect which groups of animals we are comparing. These graphs help reveal that possibility, although they cannot tell us what causes the differences or whether the differences between the groups’ size comparisons are more than we might expect by chance.

Leaving us to wonder, what exactly is sleep for anyway?


**SCIENCE VERSION**

Body Mass and Sleep Duration Across and Within Mammalian Orders

Sleep duration varies considerably across mammalian species. Proposed physiological functions of sleep include conserving energy and supporting learning and memory. However, the benefits of sleep do not necessarily explain why some species sleep longer than others. Ecological demands, such as time spent finding food and exposure to predators while resting, may also constrain sleep duration (Capellini et al., 2008). This raises the question of whether body mass has a consistent association with sleep duration across different groups of mammals.

To examine this relationship, we used the msleep dataset from the ggplot2 package, which contains observations for 83 mammalian species. The analysis focused on body mass in kilograms and total daily sleep duration in hours. The dataset’s sleep and body-mass measurements were compiled from Savage and West (2007). Because body mass varies widely across species, we applied a base-10 logarithmic transformation. We fitted a linear regression predicting daily sleep duration from log-transformed body mass.

```{r - sciplot1}
# Visualization 1: body size and daily sleep in mammals
# Data: ggplot2::msleep
# Source: https://ggplot2.tidyverse.org/reference/msleep.html
# Sleep and weight data: Savage & West (2007), doi:10.1073/pnas.0610080104

#install.packages(ggplot2)

library(ggplot2)

#calling msleep function from package
sleep_data <- ggplot2::msleep

# Create the scatterplot and add a fitted regression line.
sleep_plot_1 <- ggplot(sleep_data, aes(x = bodywt, y = sleep_total)) +
  #scatterplot for each row of data
  geom_point(size = 2.5, alpha = 0.7, colour = "grey30") +
  #linear regression line
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    colour = "black",
    fill = "grey80"
  ) +
  #body weight on a log scale as too much variance in weight normally
  #regression will still plot this even though it comes after
  scale_x_log10(
    breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000),
    labels = c("0.01", "0.1", "1", "10", "100", "1,000", "10,000")
  ) +
  #y scale
  #set continuous with break point 0-24, 4
  scale_y_continuous(breaks = c(0, 4, 8, 12, 16, 20, 24)) +
  #range
  coord_cartesian(ylim = c(0, 24)) +
  #add labels to full plot
  labs(
    title = "Total Mammal Sleep (hrs) vs Body Weight (kgs)",
        x = "Body weight (kg; log scale)",
    y = "Total sleep (hours per day)",
    caption = paste(
      "Line: linear regression; shaded band: 95% confidence interval.",
      "Source: ggplot2::msleep (Savage & West, 2007).",
      sep = "\n"
    )
  ) +
  
  theme_classic(base_size = 12)

print(sleep_plot_1)
```

Figure 1 shows a negative association across the 83 species: smaller mammals generally sleep longer than larger mammals. The fitted line summarizes this relationship, while the shaded band represents a 95% confidence interval around the estimated mean sleep duration. Although this overall pattern may seem surprising, it could partly reflect differences between groups of mammals. Does the same negative association appear when we compare species within groups of more closely related mammals?

To explore whether the overall relationship also appears within groups of related mammals, we repeated the plot separately for each mammalian order with at least six observations. This selection retained four orders: Rodentia (rodents; 22 species), Carnivora (carnivorans; 12 species), Primates (12 species), and Artiodactyla (even-toed ungulates; 6 species). The six-observation threshold was a practical choice to exclude groups with very few species. Together, these orders account for 52 of the original 83 observations. We fitted a separate regression line for each order, using the same log-transformed body mass and axis ranges as in Figure 1.

```{r - sciplot2}
#| message: false
#count the types of animal in the 'orders' (e.g. rodent, primate)
#then only take ones with greater than 6 animals
#then make new column with the n of the animal beside the order
library("dplyr")

sleep_by_order <- sleep_data |> 
  add_count(order, name = "n") |>
  filter(n >= 6) |>
  mutate(order_label = paste0(order, " (n = ", n, ")"))

#then repeat plotting process from step 1 but now broken down by the types of mammal
#using 

sleep_plot_2 <- ggplot(
  sleep_by_order,
  aes(x = bodywt, y = sleep_total)
) +
  geom_point(size = 2.5, alpha = 0.7, colour = "grey30") +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = TRUE,
    level = 0.95,
    colour = "black",
    fill = "grey80"
  ) +
  #making a unique panel for each order
  #scales 'fixed' to compare across panels
  facet_wrap(~ order_label, ncol = 2, scales = "fixed") +
  scale_x_log10(
    breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000),
    labels = c("0.01", "0.1", "1", "10", "100", "1,000", "10,000")
  ) +
  scale_y_continuous(breaks = seq(0, 24, 4)) +
  coord_cartesian(
    xlim = range(sleep_data$bodywt),
    ylim = c(0, 24)
  ) +
  labs(
    title = "The size-sleep relationship varies across groups",
    subtitle = paste(
      nrow(sleep_by_order), "of", nrow(sleep_data),
      "mammals; orders with at least 6 observations"
    ),
    x = "Body weight (kg; log scale)",
    y = "Total sleep (hours per day)",
    caption = paste(
      "Separate linear fits; shaded bands: 95% confidence intervals.",
      "Source: ggplot2::msleep (Savage & West, 2007).",
      sep = "\n"
    )
  ) +
  theme_classic(base_size = 12)

# Display the plot, including when this entire script is sourced.
print(sleep_plot_2)
```

Figure 2 shows that the overall negative association is not consistent across these groups. Among rodents and carnivorans, the fitted lines are relatively flat, suggesting little association between body mass and sleep duration within these samples. Primates show a downward relationship, with larger species generally sleeping less. The fitted line for even-toed ungulates also slopes downward, but the small sample and wide confidence band make this estimate uncertain.

These patterns suggest that the overall association may partly reflect differences between groups of mammals. However, the plots alone do not establish that the groups’ relationships differ statistically or explain what causes the differences in sleep duration.

These findings warrant a deeper investigation into the individual factors contributing to sleep patterns among orders of mammals.

**References**

Capellini, I., Barton, R. A., McNamara, P., Preston, B. T., & Nunn, C. L. (2008). Phylogenetic analysis of the ecology and evolution of mammalian     sleep. Evolution, 62(7), 1764–1776. https://doi.org/10.1111/j.1558-5646.2008.00392.x 

Savage, V. M., & West, G. B. (2007). A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of   Sciences of the United States of America, 104(3), 1051–1056. https://doi.org/10.1073/pnas.0610080104